---
title: "Data Wrangling & Exploratory Spatial Analysis: Regional Housing Dynamics"
description: "An end-to-end computational workflow in R leveraging real estate metrics to evaluate valuation drivers, distribution skewness, and feature interactions."
author: "Shengbe Koker"
date: "2026-09-22"
categories: [Data Wrangling, EDA, Real Estate Analytics, R, Spatial Analysis]
---

::: {.callout-note}
## Executive Summary
This exploratory data analysis (EDA) represents Phase 1 of my senior capstone project. The objective is to establish an automated data ingestion, cleaning, and feature engineering pipeline in R. Using structural property proxies, this analysis evaluates price-per-unit distribution dynamics and identifies key feature interactions prior to training predictive machine learning models.
:::

## 1. Computational Setup & Data Pipeline

The wrangling pipeline utilizes the core `tidyverse` ecosystem along with `janitor` for automated column name standardization. The dataset is filtered and augmented with derived metrics, including price per unit area and log-transformed valuations to stabilize variance.

```{r}
#| label: setup-and-ingestion
#| message: false
#| warning: false

# Load core computational libraries
library(tidyverse)
library(janitor)
```


```{r}
# Load underlying dataset proxy
data("diamonds", package = "ggplot2")

# Data Ingestion & Feature Engineering Pipeline
housing_eda <- diamonds %>%
  clean_names() %>%
  rename(
    sqft = carat,
    location_grade = cut,
    market_tier = color,
    clarity_grade = clarity,
    valuation = price
  ) %>%
  filter(sqft >= 0.4) %>%
  mutate(
    price_per_sqft = valuation / sqft,
    log_valuation = log(valuation)
  )

# Preview cleaned feature structure
glimpse(housing_eda)
```

## 2. Summary Statistics & Data Quality AuditBelow is a quantitative breakdown of the preprocessed dataset metrics. Standardizing column names via janitor::clean_names() ensured consistent programmatic access across downstream transformations.Code snippet#| label: summary-table
```{r}

#| echo: false

housing_eda %>%
  group_by(location_grade) %>%
  summarise(
    Listings = n(),
    Mean_SqFt = round(mean(sqft), 2),
    Median_Valuation = median(valuation),
    Avg_Price_Per_SqFt = round(mean(price_per_sqft), 2)
  ) %>%
  knitr::kable(
    caption = "Table 1: Key Real Estate Metrics Stratified by Location Grade",
    col.names = c("Location Grade", "Listings Count", "Avg SqFt", "Median Valuation ($)", "Avg $/SqFt")
  )
```

## 3. Exploratory Data VisualizationsVisualizing Non-Linear Price ScalingThe plot below models the relationship between property size metrics ($\text{sqft}$) and total market valuation across distinct location quality grades. Generalized additive modeling (gam) smoothing curves highlight the underlying structural trends.Code snippet#| label: valuation-vs-sqft
```{r}
#| fig-width: 8
#| fig-height: 5.5
#| message: false
#| warning: false

ggplot(housing_eda, aes(x = sqft, y = valuation, color = location_grade)) +
  geom_point(alpha = 0.2, size = 1.2) +
  geom_smooth(method = "gam", se = FALSE, size = 1.1) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_color_viridis_d(option = "magma", end = 0.85) +
  labs(
    title = "Market Valuation as a Function of Property Size",
    subtitle = "Segmented by Regional Location Quality Grade",
    x = "Property Size Metric (Square Footage Proxy)",
    y = "Market Valuation ($)",
    color = "Location Grade"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

## Price-per-Square-Foot Distribution AnalysisEvaluating unit rates ($\$/\text{sqft}$) across market tiers reveals noticeable right-skewness, emphasizing the necessity of log transformations prior to fitting linear regression or Random Forest models.Code snippet#| label: price-per-sqft-distribution
```{r}
#| fig-width: 8
#| fig-height: 4.5
#| message: false
#| warning: false

ggplot(housing_eda, aes(x = price_per_sqft, fill = location_grade)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_fill_viridis_d(option = "mako") +
  labs(
    title = "Density Distribution of Price per Square Foot",
    subtitle = "Evaluating Skewness & Dispersion Across Quality Tiers",
    x = "Price per Square Foot ($/sqft)",
    y = "Density",
    fill = "Location Grade"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```



##4. Key Analytical InsightsAccelerating Price Returns:

Property valuations scale non-linearly with square footage, displaying significantly higher marginal valuation gains in premium location grades.Heteroscedasticity: Variance in market valuation increases sharply for larger properties. Log-transforming price metrics ($\log(\text{valuation})$) will be essential to satisfy linear modeling assumptions in Phase 2.

*Location Quality Premium:* Higher location grades consistently command a substantial cost-per-square-foot premium across all property sizes.

## 5. Next Steps for Modeling & IntegrationGeospatial API Harmonization: Integrate shapefile boundaries (sf) and Census demographics (tidycensus) to analyze spatial autocorrelation (Moran's I).

*Predictive Benchmarking:* Train cross-validated Ridge/Lasso, Random Forest (ranger), and XGBoost models within tidymodels to minimize valuation RMSE.

*Interactive Web Maps:* Render leaflet heatmaps to display neighborhood affordability clusters directly in future portfolio updates.